---
title: "Contrasts, comparisons, and plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrasts, comparisons, and plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4.8)
```

Use this workflow after the main PheWAS runs successfully. It answers two
different questions:

| Need | Function |
|---|---|
| Compare a base and adjusted model on exactly the same participants | `run_phewas_contrast()` |
| Summarize agreement between two completed PheWAS result sets | `compare_phewas()` |

The examples use synthetic data.

```{r data}
library(phewasFlow)

example <- phewas_example_data(n = 300)
data <- example$data
eligibility <- list(
  min_n = 100L, min_cases = 10L, min_controls = 10L,
  min_outcome_levels = 3L
)

make_spec <- function(id, covariates) {
  phewas_spec(
    analysis_id = id,
    anchor = "pgs",
    direction = "phenotypes_as_outcomes",
    phenotypes = example$forward_metadata,
    covariates = covariates,
    anchor_type = "numeric",
    anchor_transform = "zscore",
    eligibility = eligibility,
    fdr_threshold = 0.05
  )
}

base_spec <- make_spec("base", "age")
adjusted_spec <- make_spec("adjusted", c("age", "sex"))
```

## Matched nested models

`run_phewas_contrast()` forms one complete-case set per phenotype and fits
both models on exactly those rows. Use it when the scientific question is how
an added covariate changes an association. This avoids mistaking a change in
the analyzed participants for a change in the model.

```{r contrast}
contrast <- run_phewas_contrast(data, base_spec, adjusted_spec)
contrast$associations[, .(
  phenotype, base_estimate, adjusted_estimate, estimate_change,
  delta_r_squared, partial_r_squared, status
)]
```

Delta and partial R-squared are reported only for continuous nested models.

## Aggregate-result comparison

Use `compare_phewas()` to describe two completed runs. This does not refit the
models or force them to use the same participants, so inspect the sample counts
before interpreting differences.

```{r comparison}
base <- run_phewas(data, base_spec)
adjusted <- run_phewas(data, adjusted_spec)
comparison <- compare_phewas(base, adjusted)
comparison$metrics
```

The comparison includes phenotypes that succeeded in both runs. Effect
correlation, slope, and RMSE are reported only when the outcome direction,
units, reference groups, and effect measures can be compared.

## Plots

Plot functions return ggplot objects. Writing files is a separate step.

```{r manhattan}
manhattan <- plot_phewas_manhattan(
  base,
  top_label_n = 3,
  label_mode = "description",
  title = "Base-model PheWAS"
)
manhattan
```

Phenotypes are grouped using `group` and colored using `color` from the
phenotype metadata. Upward and downward triangles show effect direction. Use
`label` when specific phenotypes must be named, or `top_label_n` to label the
strongest results.

The Manhattan and volcano views use the same ordered phenotype-group legend,
colors, effect-direction symbols, y-axis styling, and annotation style. Their
main difference is the x-axis: Manhattan orders phenotypes by group, while the
volcano plot shows fitted effects. This shared legend is the default. To put
group names on the Manhattan x-axis instead, use
`group_display = "x_axis"`. Use `group_display = "auto"` when you want compact
group sets on the axis and crowded sets in the legend. For command-line runs,
put the same value in `plot.group_display` in the YAML file.
With an x-axis layout, `group_label_rows` controls how many rows are available
and `group_wrap_width` controls line wrapping.

For a very large `-log10(p)` value, use
`axis_break_mode = "on"` with an `axis_break_top` chosen for the displayed
range.

Bonferroni is the default displayed threshold. Use `significance = "bh"` or
`significance = "none"` when that matches the planned figure.

The volcano view puts the fitted coefficient on the x-axis and `-log10(p)` on
the y-axis:

```{r volcano}
volcano <- plot_phewas_volcano(
  base,
  top_label_n = 3,
  label_mode = "description",
  title = "Base-model PheWAS"
)
volcano
```

The fitted coefficient remains on its model or link scale. It is not the
transformed `native_effect` such as an odds ratio. If the result contains
different effect measures, the volcano plot gives each measure its own panel.
Do not compare coefficient magnitudes across panels or across variables with
different units.

Save each plot as both a PNG for review and a vector PDF for publication or
post-editing. `save_phewas_plot()` accepts a vector of output paths:

```{r save-manhattan, eval=FALSE}
save_phewas_plot(
  manhattan,
  c("results/base-phewas.png", "results/base-phewas.pdf"),
  width = 12,
  height = 7,
  dpi = 300
)

save_phewas_plot(
  volcano,
  c("results/base-phewas-volcano.png", "results/base-phewas-volcano.pdf"),
  width = 12,
  height = 7,
  dpi = 300
)
```

```{r concordance}
plot_phewas_concordance(comparison)
```

```{r overlap}
plot_phewas_category_overlap(comparison)
```

```{r forest}
plot_phewas_forest(
  list(base, adjusted),
  phenotype_ids = example$forward_metadata$phenotype,
  run_labels = c("Base", "Adjusted")
)
```

Requested labels, highlights, or forest-plot phenotypes must exist among the
successful results. Axis limits also stop with an error if they would hide a
data point. Use `audit_file` when you also want a JSON record of the plotted
selections.
