---
title: "Scale a successful PheWAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scale a successful PheWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Scale an analysis only after it succeeds on simulated data and on a small,
representative subset of the intended phenotypes. Sharding changes where the
models run; it does not change the model formula, eligible participants, or
multiple-testing family.

Before scaling, confirm that:

- input validation succeeds;
- the anchor, covariates, references, and phenotype coding match the analysis
  plan;
- representative binary, continuous, count, or ordinal models have the
  expected sample counts and effect direction; and
- every `skipped` or `error` result in the small run is understood.

## Local parallel execution

The examples below create their own data and analysis specification.

```{r setup-example}
library(phewasFlow)

example <- phewas_example_data(n = 300)
data <- example$data
spec <- phewas_spec(
  analysis_id = "parallel_example",
  anchor = "pgs",
  direction = "phenotypes_as_outcomes",
  phenotypes = example$forward_metadata,
  covariates = c("age", "sex"),
  anchor_type = "numeric",
  anchor_transform = "zscore",
  eligibility = list(
    min_n = 100L,
    min_cases = 10L,
    min_controls = 10L,
    min_outcome_levels = 3L
  ),
  fdr_threshold = 0.05
)
```

Use the sequential backend first. On a workstation, the PSOCK backend can fit
phenotypes in parallel:

```{r local, eval=FALSE}
sequential <- run_phewas(data, spec, backend = "sequential")
parallel <- run_phewas(data, spec, backend = "psock", workers = 4)
```

Both calls return rows in the same phenotype order. Start with a small worker
count and increase it only after checking memory use.

## File-based sharding

For a restartable run, put the analysis-ready participant table, phenotype
metadata, and model settings in one YAML configuration. The runnable demo in
`inst/examples/real-data/` provides a filled example. The general reference is
`inst/examples/config/phewasflow.yml`.

The command-line workflow has five stages:

```{bash, eval=FALSE}
phewasflow validate --config analysis.yml
phewasflow manifest --config analysis.yml
phewasflow run-shard --config analysis.yml --shard-id 1
phewasflow combine --config analysis.yml
phewasflow plot --config analysis.yml
```

`validate` checks the configuration and input data. `manifest` records the
planned phenotype order, shard assignment, input hashes, model fingerprint,
and package version. Run every shard ID from `1` through
`execution.shards`; they may run concurrently.

Each shard writes one atomic RDS artifact under
`<output.directory>/.phewasflow/shards/`. The shard results do not receive
separate FDR correction. After every shard is present, `combine` verifies the
exact set and applies BH and Bonferroni correction once across all successful
models.

## Submit a SLURM array

The package includes `phewasflow_array.sbatch` and
`phewasflow_finish.sbatch` in `inst/examples/slurm/`. The commands below
assume that the installed R environment is active and the package's `exec/`
directory is in `PATH`:

```{bash, eval=FALSE}
export PATH="$PWD/exec:$PATH"
export PHEWASFLOW_CONFIG=analysis.yml

phewasflow validate --config "$PHEWASFLOW_CONFIG"
phewasflow manifest --config "$PHEWASFLOW_CONFIG"
```

Set `N_SHARDS` to the `execution.shards` value in the YAML. The submitted
array range must match it exactly.

```{bash, eval=FALSE}
N_SHARDS=10
mkdir -p logs

ARRAY_JOB_ID=$(sbatch --parsable \
  --array="1-${N_SHARDS}" \
  --output="logs/%x-%A_%a.out" \
  --error="logs/%x-%A_%a.err" \
  --export=ALL,PHEWASFLOW_CONFIG="$PHEWASFLOW_CONFIG" \
  inst/examples/slurm/phewasflow_array.sbatch)

sbatch \
  --dependency="afterok:${ARRAY_JOB_ID}" \
  --output="logs/%x-%j.out" \
  --error="logs/%x-%j.err" \
  --export=ALL,PHEWASFLOW_CONFIG="$PHEWASFLOW_CONFIG" \
  inst/examples/slurm/phewasflow_finish.sbatch
```

The array task ID selects the shard. The `afterok` dependency starts the
finish job only when every task exits successfully. Adapt the CPU, memory, and
time directives in the templates to the data size, model types, and local
scheduler policy.

Every array task reads the analysis inputs, so memory should be sized for the
full participant table, not just the phenotypes assigned to one shard. More
shards create shorter tasks and more scheduling overhead; they do not reduce
the memory needed to load the inputs. Avoid combining a large worker count
inside each task with a large array until resource use has been measured.

## Restart without mixing runs

An interrupted run does not need to start over. Resubmit the same array with
the same YAML. A valid existing shard is fingerprint-checked and reused.
Missing shards run normally, and mismatched artifacts stop with a specific
error.

`combine` also checks for missing or unexpected shard files and refuses to
mix results from different specifications. If existing combined results match
the validated shards, they are reused and the portable TSV is recreated.

Use a new `analysis_id` and output directory whenever any scientific or
computational input changes, including:

- participant data or phenotype metadata;
- the anchor, covariates, exclusions, references, or thresholds;
- the shard count; or
- the installed package version.

Keeping each run in its own output directory is simpler and safer than trying
to replace selected artifacts from an older run.

## Review the completed run

The configured output directory contains:

```text
results.rds
results.tsv
phewas.png
phewas.pdf
phewas-volcano.png
phewas-volcano.pdf
.phewasflow/
  manifest.tsv
  shards/
```

`plot.output` sets the plot stem. One `phewasflow plot` command writes the
Manhattan and volcano plots in both PNG and PDF; the names above result from a
stem of `phewas`. The configured significance line is Bonferroni by default.
Review the combined result rather than individual shard files:

```{r review, eval=FALSE}
result <- read_phewas_bundle("phewasflow-output")

table(result$status, useNA = "ifany")
subset(
  result,
  status != "ok",
  select = c(
    phenotype, status, reason_code, message, warnings,
    n_complete, n_event, n_nonevent
  )
)
```

`ok` rows form the multiple-testing family. A `skipped` row records an
eligibility or estimability decision; an `error` row records a validation,
fitting, or convergence failure. Check these statuses, sample counts,
`testing_family_size`, effect estimates, and adjusted p-values before
interpreting or sharing the analysis.

For a copyable submission sequence and log layout, see
`inst/examples/slurm/README.md`.
