---
title: "Using PheWAS with phewasFlow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using PheWAS with phewasFlow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  eval = requireNamespace("PheWAS", quietly = TRUE)
)
```

## Why this package exists

`phewasFlow` packages the work around repeated association scans: declaring an
analysis, checking inputs, restarting cluster tasks, auditing individual
results, and comparing covariate adjustments on the same participants. Its
inputs are already constructed phenotypes.

The [PheWAS package](https://github.com/PheWAS/PheWAS) provides a broader toolkit
for clinical-code mapping, phenotype construction, regression, plotting, and
meta-analysis. The packages share capabilities such as Firth regression and
parallel execution; phewasFlow focuses on the workflow around prepared data.

| Task | Role of phewasFlow |
|---|---|
| Construct phenotypes from clinical codes | Accept prepared phenotypes from PheWAS or another upstream workflow |
| Run Firth or linear regression | Provide explicit specifications and detailed result records around existing model engines; these models are shared capabilities |
| Run on a cluster and resume | Fingerprint inputs and specifications, reuse matching completed shards, reject incomplete or incompatible shard sets |
| Correct a sharded scan | Apply BH and Bonferroni once across successful models in the combined testing family |
| Compare additional adjustment | Fit both specifications on the same complete-case rows for each phenotype |
| Model count and ordered outcomes | Provide negative-binomial models with optional log-offsets and proportional-odds ordinal models |
| Retain very small p-values | Calculate upper tails in log space and preserve log-scale values through correction and reporting |

The packages can work together: construct phenotypes with PheWAS, then use
phewasFlow's own model wrappers for repeated scans and matched model comparisons.

## Retain very small p-values

Calculating a small tail probability by subtracting from one can lose precision
or round to zero. phewasFlow evaluates upper-tail probabilities directly in log
space and keeps that representation through BH and Bonferroni correction,
preserving very small p-values for reporting and plotting.

Use `log_p`, `p_value_text`, and `neg_log10_p` when probabilities are too small
for the ordinary numeric `p_value` field. Adjusted results also have text and
negative-log representations.

## Run the synthetic bridge

PheWAS is an optional suggested dependency. Install the pinned upstream
revision to run this article's example:

```{r install-phewas, eval=FALSE}
remotes::install_github(
  "PheWAS/PheWAS@55dd1c24e228851922400cfba8d7db474565ccc7",
  dependencies = NA, upgrade = "never"
)
```

The installed example creates 400 simulated participants and dated ICD9CM
records. It calls `PheWAS::createPhenotypes()` with a two-record case threshold,
sex information, and the full eligible sample list, then fits a synthetic PRS
scan with phewasFlow. No clinical findings are represented.

```{r run-bridge}
source(system.file(
  "examples", "phewas", "use-phewas-phenotypes.R", package = "phewasFlow"
))
bridge <- run_phewas_bridge()
table(bridge$result$status)
bridge$result[bridge$result$status == "ok",
              c("phenotype", "description", "status", "n_complete", "n_event")]
```

PheWAS's exclusion mappings also create phenotype columns with no observed
cases in this tiny simulation. phewasFlow retains those as `skipped` rows with
reason `insufficient_cases`. They are not part of the successful testing family.

## Preserve the phenotype definitions

For your own records, the upstream step has this form. The first four columns
of `records` are participant ID, vocabulary, code, and date. `sample$id` must
represent the population eligible for the intended analysis.

```{r construct-phenotypes, eval=FALSE}
phenotypes <- PheWAS::createPhenotypes(
  records,
  min.code.count = 2L,
  id.sex = covariates[c("id", "sex")],
  full.population.ids = sample$id
)
```

Review the mapping version, observation window, case threshold, control
definitions, and sex restrictions for your study. The example demonstrates
software interoperability; it does not establish that these definitions are
appropriate for your data.

With a case threshold, PheWAS returns logical `TRUE` for cases, `FALSE` for
controls, and `NA` for excluded or uncertain observations. Preserve `NA`.
Do not fill it with `FALSE` or zero. Keep exact phecode column names when
reading or converting data frames; use `check.names = FALSE` with base readers.

```{r bridge-metadata}
library(PheWAS)
phenotypes <- bridge$phenotypes
phenotype_ids <- setdiff(names(phenotypes), "id")
metadata <- as.data.frame(PheWAS::addPhecodeInfo(
  data.frame(phenotype = phenotype_ids),
  groupnums = TRUE, groupcolors = TRUE
))
stopifnot(!anyDuplicated(metadata$phenotype))
stopifnot(setequal(metadata$phenotype, phenotype_ids))
metadata <- metadata[match(phenotype_ids, metadata$phenotype),
                     c("phenotype", "description", "group", "groupnum", "color")]
metadata$variable_type <- "binary"
metadata$outcome_type <- "binary"
metadata$reference <- "FALSE"
head(metadata, 6L)
```

The reference is the character string `"FALSE"` because the participant columns
are logical. The default `"0"` applies to numeric 0/1 coding. Alternatively,
convert logical columns with `as.integer()` (which preserves `NA`) and declare
`"0"` explicitly. Do not change the reference without changing the encoding.

Attach PheWAS with `library(PheWAS)` before calling its metadata helper; the
pinned version looks up its `pheinfo` dataset on the search path.
`addPhecodeInfo()` uses an inner join. The checks above prevent an unknown code
from disappearing silently. For custom or unmapped phenotypes, supply their
description, group, group order, and color explicitly.

Join the phenotype table to the declared sample and other participant tables:

```{r join-tables, eval=FALSE}
analysis <- phewasFlow::assemble_phewas_data(
  sample,
  phenotypes = phenotypes,
  anchors = anchors,
  covariates = covariates,
  id = "id"
)
print(attr(analysis, "join_report"))
```

Keep categorical covariates as factors with declared reference levels. The
sample and each input table must have unique, nonmissing IDs of the same type.
The full runnable example creates a `phewas_spec()` and calls `run_phewas()`;
inspect `bridge$spec` and `bridge$analysis` to see the exact inputs. For real
data configuration and cluster execution, continue with
[the data workflow](real-data-workflow.html) and [large runs](large-runs.html).

## Compare numerical results and runtime

The installed comparison script generates shared binary and continuous
outcomes, then compares PheWAS and phewasFlow on the same data, covariates,
references, and eligibility thresholds. Run it from a source checkout:

```{bash compare-packages, eval=FALSE}
Rscript inst/examples/phewas/compare-packages.R output/phewas-comparison
```

The default workload has 1,000 participants and four phenotypes of each type.
Each package uses one worker, one unmeasured warmup, and three measured repeats
with alternating execution order. Both receive missing anchor, covariate, and
phenotype values. The binary comparison uses `phewas_ext(method = "logistf")`;
the continuous comparison uses `phewas()`.

The script checks coefficients, standard errors, natural-log p-values, and
sample counts, including case/control counts for binary outcomes. It records
parameters, package versions, the installed upstream revision, individual
timings, and a summary. A numerical mismatch fails the command after writing
the comparison tables. No participant data or fitted models are saved.

Timings include scan validation and result assembly, but exclude simulation,
specification construction, plotting, disk I/O, and multiple-testing correction.
These small forward-scan workloads cover binary and continuous outcomes;
timings depend on the data, hardware, and package versions. The
[example README](https://github.com/FritscheLab/phewasFlow/tree/main/inst/examples/phewas)
documents the tolerances, commands, outputs, and limitations in full.

## Recorded comparison

The following synthetic runs were measured on September 12, 2026, using R
4.6.1 on ARM64 macOS, phewasFlow 0.1.0, PheWAS 0.99.6-1 at revision
`55dd1c24e228851922400cfba8d7db474565ccc7`, and logistf 1.26.1. Both used seed
20260912 and the methods described above. The timings are median elapsed
seconds per scan of one outcome type; they are specific to this machine and
workload, and the shortest measurements are especially sensitive to overhead.

| Participants | Phenotypes per type | Repeats | Outcome | PheWAS seconds | phewasFlow seconds |
|---:|---:|---:|---|---:|---:|
| 1,000 | 4 | 3 | Binary | 0.022 | 0.031 |
| 1,000 | 4 | 3 | Continuous | 0.005 | 0.011 |
| 5,000 | 10 | 5 | Binary | 0.179 | 0.283 |
| 5,000 | 10 | 5 | Continuous | 0.023 | 0.037 |

PheWAS was faster in these runs. phewasFlow's validation, detailed result
construction, and additional Firth model fits contribute to its runtime.
The log-scale probability calculation itself does not require extra fits.

Sample counts matched, and coefficients and standard errors agreed to numerical
precision in both runs. All eight associations met the comparison tolerances
in the smaller run. In the larger run, five of 20 associations had very small
p-values that differed beyond the specified tolerance:

| Synthetic outcome | PheWAS p-value | phewasFlow p-value |
|---|---:|---:|
| `binary_003` | `5.66214e-15` | `5.64294e-15` |
| `binary_004` | 0 | `3.37208e-20` |
| `binary_006` | `5.55112e-16` | `5.82611e-16` |
| `binary_007` | `1.11022e-16` | `1.60862e-16` |
| `binary_008` | 0 | `7.13610e-36` |

Values are rounded for display. In the tested `logistf` version, the calculation
`1 - pchisq(...)` can lose precision or return zero for very small p-values,
explaining both the nonzero and zero differences above. phewasFlow evaluates
the upper tail directly in log space; an independent high-precision check
confirmed these probabilities for the same fitted statistics. The association
test is unchanged.
