---
title: "Run a PheWAS with your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run a PheWAS with your data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This guide starts with an analysis-ready participant table and ends with a
result table plus Manhattan and volcano plots. An analysis-ready phenotype
already has its case, control, timing, and exclusion rules applied. For
example, a binary phecode column should contain `1` for cases, `0` for
controls, and `NA` when the phenotype is unavailable.

## Choose the analysis direction

`phewasFlow` supports two common questions:

| Question | `direction` | Fixed variable (`anchor`) | Scanned phenotypes |
|---|---|---|---|
| Is one PRS, exposure, or biomarker associated with many outcomes? | `phenotypes_as_outcomes` | Predictor | Outcomes |
| Are many diagnoses or traits associated with one clinical endpoint? | `phenotypes_as_predictors` | Outcome | Predictors |

One configuration is one multiple-testing family. Use a separate
`analysis_id` and output directory for each PRS, exposure, or fixed clinical
endpoint.

## Prepare the participant table

The model input is a data frame with one row per participant. It must contain
the anchor, all covariates, every phenotype named in the phenotype metadata,
and any declared offset or exclusion columns. A participant ID is recommended
for joining and checking records, but it is not included in the regression
model.

For a PRS-to-phecode analysis, the table might look like this:

| IID | prs | age | sex | PH_001 | PH_002 |
|---|---:|---:|---|---:|---:|
| P001 | -0.42 | 54 | female | 0 | 1 |
| P002 | 0.18 | 61 | male | 1 | 0 |
| P003 | 0.73 | 47 | female | NA | 0 |

Use these representations:

| Variable | Representation |
|---|---|
| Numeric predictor or continuous outcome | Finite numeric values |
| Binary variable | No more than two observed values; identify the intended reference value |
| Count outcome | Nonnegative integers |
| Ordinal variable | Values matching the declared ordered levels exactly |
| Covariate | Numeric, logical, or factor |
| Count offset | The finite log of a positive exposure value |

Convert missing-value codes such as `-9` or `999` to `NA`. Do not convert a
missing phenotype to a control. Missing anchor or covariate values remove the
participant from the common base cohort; phenotype missingness is handled
separately for each association.

RDS is the easiest input format when a covariate is categorical because it
preserves factor levels. With CSV or TSV input, encode a categorical covariate
as K-1 numeric indicator columns and document the omitted reference group.

If the variables are stored in separate files, join them to an explicit sample
list before saving the analysis table:

```{r assemble-data, eval=FALSE}
library(data.table)
library(phewasFlow)

sample <- fread("source/sample.tsv", colClasses = c(IID = "character"))
phenotypes <- fread(
  "source/phenotypes.tsv", colClasses = c(IID = "character")
)
anchors <- fread("source/anchors.tsv", colClasses = c(IID = "character"))
covariates <- fread(
  "source/covariates.tsv", colClasses = c(IID = "character")
)

covariates[, sex := factor(sex, levels = c("female", "male"))]

analysis <- assemble_phewas_data(
  sample = sample,
  phenotypes = phenotypes,
  anchors = anchors,
  covariates = covariates,
  id = "IID"
)

print(attr(analysis, "join_report"))
stopifnot(!anyDuplicated(analysis$IID))

dir.create("inputs", showWarnings = FALSE)
saveRDS(analysis, "inputs/analysis.rds", version = 3)
```

`assemble_phewas_data()` requires one nonmissing, unique ID per source table.
Its join report shows records that did not match the sample list. Review those
counts before fitting models.

## Prepare phenotype metadata

The metadata file has one row per scanned phenotype. `phenotype` must exactly
match a column in the participant table.

These columns are always required:

| Column | Meaning |
|---|---|
| `phenotype` | Phenotype identifier and participant-table column name |
| `description` | Readable phenotype label |
| `group` | Clinical or phenotype group shown in the plot |
| `groupnum` | Numeric group order |
| `color` | Valid R color for the group |

`variable_type` may be `numeric`, `binary`, `count`, or `ordinal`. If the
column or a row value is missing, `NA`, or blank, that scanned phenotype
defaults to `binary`.

Every row in one `group` must use the same `groupnum` and `color`, and each
group must have its own `groupnum`.

When phenotypes are outcomes, `outcome_type` is still required for every row;
it is not inferred from `variable_type`:

| `variable_type` | `outcome_type` | Model |
|---|---|---|
| `numeric` | `continuous` | Linear regression |
| `binary` | `binary` | Firth logistic regression |
| `count` | `count` | Negative-binomial regression |
| `ordinal` | `ordinal` | Proportional-odds ordinal regression |

Some variable types need additional columns:

| Column | When needed | Value |
|---|---|---|
| `reference` | Binary phenotype | Reference value; missing, `NA`, or blank defaults to character `0` |
| `levels` | Ordinal phenotype | Ordered values separated by `|`, such as `mild|moderate|severe` |
| `scores` | Optional for an ordinal predictor | Increasing scores separated by `|` |
| `offset` | Optional for a count outcome | Name of an already-log-transformed offset column |

Example forward metadata:

```text
phenotype  description          group           groupnum  color    variable_type  outcome_type  reference  levels  scores  offset
PH_001     Type 2 diabetes      Endocrine       1         #2563EB  binary         binary        0          NA      NA      NA
PH_002     C-reactive protein   Measurements    2         #059669  numeric        continuous    NA         NA      NA      NA
```

Use tab-separated fields in the actual file. For a
`phenotypes_as_predictors` analysis, omit `outcome_type`; the fixed outcome type
is declared in the YAML configuration. Scanned predictors may be numeric,
binary, or ordinal. Declare `variable_type` and `reference` explicitly whenever
the binary defaults do not describe a scanned phenotype.

## Configure the analysis

Copy `analysis-forward.yml` from `inst/examples/real-data/` and edit its paths
and variable names. Relative paths are resolved from the directory containing
the YAML file.

This is a one-shard PRS-to-phenome configuration:

```yaml
schema_version: 1
inputs:
  data: inputs/analysis.rds
  phenotype_metadata: phenotype-metadata-outcomes.tsv
analysis:
  analysis_id: my_prs_to_phenome_v1
  direction: phenotypes_as_outcomes
  anchor: prs
  anchor_type: numeric
  anchor_transform: zscore
  covariates: [age, sex]
  fdr_threshold: 0.05
  eligibility:
    min_n: 100
    min_cases: 10
    min_controls: 10
    min_outcome_levels: 3
execution:
  shards: 1
  backend: sequential
  workers: 1
output:
  directory: phewasflow-output/prs-to-phenome
plot:
  # Creates phewas.png, phewas.pdf, phewas-volcano.png, and
  # phewas-volcano.pdf.
  output: phewasflow-output/prs-to-phenome/phewas.png
  significance: bonferroni
  group_display: legend
  label: []
  highlight: []
  width: 10
  height: 6
  dpi: 300
```

`anchor_transform: zscore` reports associations per standard deviation of a
numeric predictor. The mean and standard deviation are calculated once in the
post-exclusion cohort with complete anchor and covariate data.

`min_cases` and `min_controls` set the minimum non-reference and reference
counts for each scanned binary phenotype. The same thresholds therefore apply
when binary phenotypes are outcomes and when they are predictors. If the fixed
outcome is binary, its event and nonevent counts must also meet these minima.

For a phenotype-to-endpoint analysis, start from `analysis-reverse.yml`. The
important declarations are:

```yaml
inputs:
  phenotype_metadata: phenotype-metadata-predictors.tsv
analysis:
  analysis_id: diagnoses_to_endpoint_v1
  direction: phenotypes_as_predictors
  anchor: endpoint_continuous
  anchor_type: numeric
  outcome_type: continuous
  anchor_transform: none
```

Keep the covariates, eligibility thresholds, execution, output, and plot
sections from the forward example. A fixed binary anchor always requires an
explicit `anchor_reference`; the scanned-phenotype default does not apply. For
a binary fixed outcome, set `outcome_type: binary` and provide
`anchor_reference`; `anchor_type: binary` may also be supplied explicitly. An
ordinal fixed outcome requires ordered `anchor_levels`. A count fixed outcome
may name an already-log-transformed `offset` column.

## Run locally

Install `phewasFlow`, then run the provided project runner from the directory
containing the configuration:

```{bash, eval=FALSE}
Rscript run-local.R analysis-forward.yml
```

The runner validates the configuration and columns, runs every configured
shard, combines the association results, applies BH and Bonferroni correction,
and creates Manhattan and volcano plots in PNG and PDF. Validation cannot
detect every value problem or model convergence issue, so review association
statuses after the run.

To verify the setup with the included example first:

```{bash, eval=FALSE}
Rscript inst/examples/real-data/create-demo-project.R my-phewas
cd my-phewas
Rscript run-local.R analysis-forward.yml
```

The generated participant data are simulated. Replace `inputs/analysis.rds`,
the phenotype metadata, and the YAML settings before using the project for an
analysis.

## Review the results

When at least one model has a finite p-value, the output directory contains:

```text
results.rds
results.tsv
phewas.png
phewas.pdf
phewas-volcano.png
phewas-volcano.pdf
```

`plot.output` supplies the filename stem, even when it ends in `.png` or
`.pdf`. The plot command always writes both formats. Bonferroni is the default
displayed threshold; set `plot.significance` to `bh` or `none` only when that
matches the planned figure.

The volcano x-axis is the fitted coefficient on the model or link scale. It is
not the transformed `native_effect`. Different effect measures are separated
into panels, and coefficient magnitudes should not be compared across panels
or across variables measured in different units.

The RDS file preserves the full specification and run information. The TSV is
convenient for review in other software. Load the result and check every model
status:

```{r review-results, eval=FALSE}
library(phewasFlow)

result <- read_phewas_bundle("phewasflow-output/prs-to-phenome")
table(result$status, useNA = "ifany")

subset(
  result,
  status != "ok",
  select = c(phenotype, status, reason_code, message, warnings)
)

subset(
  result,
  status == "ok",
  select = c(
    phenotype, description, n_complete, n_event, n_nonevent,
    effective_formula, dropped_invariant_covariates,
    native_effect, native_conf_low, native_conf_high,
    p_value_text, q_value_text, fdr_significant
  )
)
```

Each requested phenotype has one of three statuses:

| Status | Meaning |
|---|---|
| `ok` | Model completed and entered multiple-testing correction |
| `skipped` | Data did not meet an eligibility or estimability requirement |
| `error` | Data validation, model fitting, or convergence failed |

Check `reason_code`, `message`, and `warnings` for every non-`ok` row. Also
review `n_base`, `n_complete`, case/control or level counts, reference values,
and `testing_family_size` against the analysis plan.

`formula` and `covariates` record the requested model. If a covariate is
constant in one phenotype's complete-case sample, it cannot contribute to
that model and is omitted automatically. `effective_formula`,
`effective_covariates`, and `dropped_invariant_covariates` record exactly what
was fitted. Other collinearity remains a `rank_deficient` skip.

`estimate` is on the fitted model's link scale. `native_effect` is easier to
interpret:

| Outcome | `native_effect` |
|---|---|
| Continuous | Regression coefficient or adjusted mean difference |
| Binary | Odds ratio |
| Count | Incidence-rate ratio |
| Ordinal | Common odds ratio |

The binary reference and ordinal level order determine the effect direction.
For very small p-values, use `p_value_text`, `q_value_text`, or
`neg_log10_p`; the ordinary numeric p-value can reach the floating-point limit.

When changing the participants, phenotypes, model, or thresholds, use a new
`analysis_id` and output directory. For larger local or cluster runs, continue
with `vignette("large-runs", package = "phewasFlow")` after this one-shard
workflow succeeds.

## Common problems

| Message or symptom | What to check |
|---|---|
| Missing declared column | Match YAML and metadata names exactly to participant-table columns |
| Character covariate | Save a factor in RDS or use K-1 numeric indicator columns |
| `reference_not_observed` | Match the declared binary reference to the encoded value |
| `undeclared_ordinal_level` | Match spelling, capitalization, and order to `levels` |
| `insufficient_cases` or `insufficient_controls` | Review phenotype coding, missingness, and eligibility thresholds |
| `rank_deficient` | Check for duplicated or collinear covariates; invariant covariates are omitted automatically |
| No plot | Confirm that at least one association has `status == "ok"` and a finite p-value |
