---
title: "PheWAS in both directions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PheWAS in both directions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Choose the scan direction from the scientific question:

| Question | Direction | Repeated model |
|---|---|---|
| Is one PRS, exposure, or biomarker associated with many phenotypes? | `phenotypes_as_outcomes` | Each phenotype is the outcome; the anchor is the predictor |
| Is each phenotype associated with one clinical endpoint? | `phenotypes_as_predictors` | The clinical endpoint is the outcome; each phenotype is a predictor |

The examples below use synthetic participants so they can run without access
to clinical data.

```{r data}
library(phewasFlow)

example <- phewas_example_data(n = 300)
data <- assemble_phewas_data(
  example$sample, example$phenotypes, example$anchors,
  example$covariates, id = "id"
)

eligibility <- list(
  min_n = 100L,
  min_cases = 10L,
  min_controls = 10L,
  min_outcome_levels = 3L
)
```

## PRS or PGS as the predictor

The phenotype metadata explicitly declares binary, continuous, count, and
ordinal outcomes. The anchor is standardized once on the common base cohort,
before phenotype-specific missingness is considered.

```{r forward}
forward_spec <- phewas_spec(
  analysis_id = "pgs_to_phenome",
  anchor = "pgs",
  direction = "phenotypes_as_outcomes",
  phenotypes = example$forward_metadata,
  covariates = c("age", "sex"),
  anchor_type = "numeric",
  anchor_transform = "zscore",
  eligibility = eligibility,
  fdr_threshold = 0.05
)

forward <- run_phewas(data, forward_spec)
forward[, .(phenotype, description, engine, estimate, native_effect, status)]
```

Binary and ordinal effects are odds ratios, count effects are incidence-rate
ratios, and continuous effects remain regression coefficients. The `estimate`
column always contains the coefficient on the model's link scale.

For a count outcome with unequal exposure time, name a column containing the
already-log-transformed exposure in that phenotype row's `offset` field. Leave
`offset` empty when no offset belongs in the model.

## Phenotypes as predictors

This example asks whether each phenotype is associated with a continuous
clinical endpoint. The endpoint stays fixed while `phewasFlow` changes the
phenotype predictor.

```{r reverse}
reverse_spec <- phewas_spec(
  analysis_id = "phenome_to_continuous_endpoint",
  anchor = "endpoint_continuous",
  direction = "phenotypes_as_predictors",
  phenotypes = example$reverse_metadata,
  covariates = c("age", "sex"),
  outcome_type = "continuous",
  anchor_type = "numeric",
  eligibility = eligibility,
  fdr_threshold = 0.05
)

reverse <- run_phewas(data, reverse_spec)
reverse[, .(phenotype, description, estimate, native_effect, status)]
```

Missing phenotype values are excluded association by association. They are
never converted to a binary reference group.

For a binary clinical endpoint, set `outcome_type = "binary"`,
`anchor_type = "binary"`, and give its reference value with
`anchor_reference`. Count and ordinal endpoints use `outcome_type = "count"`
or `"ordinal"`; ordinal endpoints also require `anchor_levels` in their true
order. Use a separate `analysis_id` and result directory for each clinical
endpoint because each one has its own multiple-testing correction.
